An elaborate data set on human gait and the effect of mechanical perturbations

### 2015-03-17 paper.tex
Made the distinction between pilot trials and the final protocol much more apparent.

% Editing notes
% - No tabs, just spaces!
% - No hanging whitespace
% - Wrap lines to 79 characters
% - Indentation should be two spaces per indent
% - Only one space after periods.
% - Don't worry about formatting details, just get the content in place.

@@

\subsection*{Participants}
%
Fifteen able bodied subjects including four females and eleven males with an
average age of $24\pm4$ years, height of $1.75\pm0.09$ m, mass of $74\pm13$ kg
participated in the study. The study was approved by the Institutional Review
Board of Cleveland State University (\# 29904-VAN-HS) and written informed
consent was obtained from all participants. The data has been anonymized with
respect to the participants' identities and a unique identification number was
assigned to each subject. A selection of the meta data collected for each
subject is shown in Table~\ref{tab:subjects}.
%
\begin{table}
  \cprotect\caption{Information about the 15 study participants. The final
    three columns provide the trial numbers associated with each nominal
    treadmill speed. The measured mass is computed from the mean total vertical
    ground reaction force just after the calibration pose event, if possible.
    If the mass is reported without an accompanying standard deviation, it is
    the subject's self-reported mass. Additional trials found in the data set
    with a subject identification number 0 are trials with no subject, i.e.
    unloaded trials that can be used for inertial compensation purposes, and
    are not shown in the table. Generated by \verb|src/subject_table.py|.}
  \centering
  \small
  \input{tables/subjects.tex}
  \label{tab:subjects}
\end{table}

@@

Early on, we discovered that the factory setup of the R-Link treadmill had a
vibration mode as low as 5\si{\hertz} that was detectable in the force
measurements, likely due to the flexible undercarriage and pitch motion
mechanism. Trials 6-8 are affected by this vibration mode. During trials 9-15
the treadmill was stabilized with wooden blocks. During the remaining trials
the treadmill was stabilized with metal supports; both in-house fabricated ones
and vendor supplied ones. These supports aimed to improve the stiffness and
frequency response of the force plate system. See the Data Limitations Section
for more details.

The acceleration of the treadmill base was measured during each trial by four
ADXL330 accelerometers placed at the four corners of the machine. These
accelerometers were intended to provide information for inertial compensation
purposes when the treadmill moved laterally or in pitch, but are extraneous for
trials greater than number 8 due to the treadmill being stabilized in those
degrees of freedom.
%
\begin{figure}
  \centering
  \includegraphics{figures/treadmill.png}
  \caption{The treadmill with coordinate system, cameras (circled in orange),
    projection screen, and safety rope. The direction of travel is in the $-z$
    direction.}
  \label{fig:treadmill}
\end{figure}

@@

hair but at the expense of accuracy.}, and a rock climbing harness. All 47
markers were applied directly to the skin at the landmarks noted in

@@

  \cprotect\caption{Descriptions of the 47 markers used in this study. The
    ``Set'' column indicates whether the marker exists in the lower and/or full
    body marker set. The label column matches the column headers in the

@@

After obtaining informed consent and a briefing by the experimentalist on the
trial protocol, the subject followed the verbal instructions of the
experimentalist and the on-screen instructions from the video display. The
protocol for a single trial was as follows:
%
\begin{enumerate}
  \item Subject stepped onto the treadmill and markers were identified with
    Cortex.
  \item The safety rope was attached loosely to the rock climbing harness such
    that no forces were acting on the subject during walking, but that the
    harness would prevent a full fall.
  \item The subject started by stepping on sides of treadmill so that feet did
    not touch the force plates and the force plate signals are zeroed. This
    corresponds to the ``Force Plate Zeroing'' event.
  \item Once notified by the video display, the subject stood in the
    calibration pose: standing straight up, looking forward, arms out by their
    sides (approximately 45 degree abduction) and the event, ``Calibration
    Pose'', was manually recorded by the operator.
  \item A countdown to the first normal walking phase was displayed. At the end
    of the countdown the event ``First Normal Walking'' was recorded and the
    treadmill ramped up to the specified speed and the subject was instructed
    to walk normally, to focus on the ``endless'' road on the display, and not
    to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begun and was recorded as ``Longitudinal Perturbation''.
  \item After 8 minutes of walking under the influence of the perturbations,
    the second normal walking phase begun and was recorded as ``Second Normal
    Walking''.
  \item After 1 minute of normal walking, a countdown was shown on the display
    and the treadmill decelerated to a stop.
  \item The subject was instructed to step off of the force plates for 10
    seconds and the ``Unloaded End'' event was recorded.
  \item The subject could then take a rest break before each additional trial.
\end{enumerate}

Trials 6-8 included a calibration pose at the start of the trial but the event
was not explicitly recorded. In those trials, the ``TreadmillPerturbation''
event marks the beginning of longitudinal perturbations and the ``Both'' event
marks the beginning of combined longitudinal and lateral perturbations. The
force plate zeroing at the end was also not explicitly recorded.

@@

In addition to the longitudinal perturbations, lateral perturbations were also
prescribed for a duration of four minutes for one subject in earlier trials
(6,7,8). Figure~\ref{fig:lateral_perturbation} shows the effect of the lateral
control of the treadmill by plotting the commanded and measured treadmill
positions. These signals were generated in a similar manner using MATLAB and
Simulink in which a Gaussian white noise block that twice integrated to obtain
the lateral position. The signal was then high-pass filtered with a
second-order Butterworth filter to eliminate drift and then saturated so that
the signal remained within the 5~\si{\centi\meter} lateral range of the
physical hardware.  The same perturbation signal was used for each of the three
trials.
%
\begin{figure}
  \centering
  \includegraphics{figures/lateral_perturbation.pdf}
  \cprotect\caption{Commanded treadmill lateral deviation (red) and the
    recorded deviation (blue) for the trials 6, 7, and 8. Generated by
    \verb|src/lateral_perturbation_plot.m|.}
  \label{fig:lateral_perturbation}
\end{figure}

@@

\subsubsection*{mocap-xxx.txt}
%
The output from the D-Flow mocap module is stored in a tab separated value
(TSV) file named \verb+mocap-xxx.txt+ where \verb+xxx+ represents the trial id
number. The file contains a number of time series. The numerical values of the
time series are provided in decimal fixed point notation with 6 decimals of
precision, e.g. \verb|123456.123456|, regardless of the units. The first line
of the file holds the header. The header includes time stamp column, frame
number column, marker position columns, force plate force/moment columns, force
plate center of pressure columns, and other analog columns. The columns are
further described below:
%
\begin{description}
  \item[TimeStamp] The monotonically increasing computer clock time when D-Flow
    receives a frame from Cortex. These are recorded approximately at
    100~\si{\hertz} sampling rate and given in seconds.
  \item[FrameNumber] Monotonically increasing positive integers that correspond
    to each frame received from Cortex.
  \item[Marker Coordinates] Any column that ends in \verb+.PosX+, \verb+.PosY+,
    or \verb+.PosZ+ are marker coordinates expressed in Cortex's Cartesian
    reference frame. The prefixes match the marker labels given in
    Table~\ref{tab:marker-labels}. These values are in meters.
  \item[Ground Reaction Loads] There are three ground reaction forces and three
    ground reaction moments recorded by each of the two force plates in Newtons
    and Newton-Meters, respectively. The prefix for these columns is either
    \verb+FP1+ or \verb+FP2+ and represents either force plate 1 (left) or 2
    (right). The suffixes are either \verb+.For[XYZ]+, \verb+.Mom[XYZ]+ for the
    forces and moments, respectively. The force plate voltages are sampled at a
    much higher frequency than the cameras, but delivered at the Cortex camera
    sample rate, \~100~\si{\hertz} through the D-Flow mocap module. A
    force/moment calibration matrix stored in Cortex converts the voltages to
    forces and moments before sending it to D-Flow. The software also computes
    the center of pressure from the forces, moments, and force plate
    dimensions. These have the same prefixes for the plate number, have the
    suffix \verb+.Cop[XYZ]+, and are given in meters.
  \item[Analog Channels] Several analog signals are recorded under column
    headers \verb+Channel[1-99].Anlg+. These correspond to analog signals
    sampled by Cortex and correspond to the 96 analog channels in the National
    Instruments USB-6255. The first twelve are the voltages from the
    force plate load cells. We also record the acceleration of 4 points on the
    treadmill base in analog channels 61-72 that were in place in case inertial
    compensation for the lateral treadmill movement was required.
\end{description}

@@

The \verb|DFlowData| class collects and stores all the raw data presented in
the previous section and applies several ``cleaning'' operations to transform
the data into a usable form. The cleaning process follows these steps:
%
\begin{enumerate}
  \item Load the meta data file into a Python dictionary.
  \item Load the D-Flow mocap module TSV file into Pandas \verb|DataFrame|.
  \item Relabel the column headers to more meaningful names if this is
    specified in the meta data.
  \item Optionally identify the missing values in the mocap marker data and
    replace them with \verb|numpy.nan|.
  \item Optionally interpolate the missing marker values and replaces them
    with interpolated estimates using a variety of interpolation methods.
  \item Load the D-Flow record module TSV file into a Pandas \verb|DataFrame|.
  \item Extract the events and create a dictionary mapping the event names in
    the meta data to the events detected in the record module file.
  \item Interially compensate the ground reaction loads based on whether the
    meta data indicates there was treadmill motion.
  \item Merge the data from the mocap module and record module into one data
    frame at the maximum common constant sample rate.
\end{enumerate}

@@

\subsection*{Data Limitations}
%
The data is provided in good faith with great attention to detail but as with
all data there are anomalies that may affect the use and interpretation of
results emanating from the data. The following list gives various notes and
warnings about the data that should be taken into account when making use of
it.
%
\begin{itemize}
  \item Be sure to read the notes in each meta data file for details about
    possible anomalies in that particular trial. Things such as marker dropout,
    ghost markers, and marker movement are the more prominent notes. Details
    about variations in the equipment on the day of the trial are also
    mentioned.
  \item The subject identification number 0 stands for "no subject" and was
    used whenever data was collected from the system with no subject on the
    treadmill, for example during the trials that were intended to be used for
    inertial compensation purposes. These trials play through the exact
    protocol as those with a human subject and the matching trials are
    indicated in the meta data. Matching unloaded trials were recorded on the
    same day as the loaded trials and is noted in the
    \verb|trial:files:compensation| section of the meta data file. See
    Table~\ref{tab:comptrials} for a list of all the compensation trials.
  \item Trials 1 and 2 were not recorded as part of this study. Those trial
    identification numbers were reserved for early data exploration from data
    collected in other studies.
  \item Subject 1 walked only at a single speed with three trial repetitions.
    This is because these trials were originally for protocol testing purposes
    but the kinematic data is valid and the trials uniquely have lateral
    perturbations, so we have included them. See the next items for more
    details about these trials.
  \item Trials 6-8 use a only the lower body marker set. The remaining trials
    are full body.
  \item Trials 6-8 use an early experimental protocol which divided the
    perturbation sections into three sections: longitudinal perturbations,
    lateral perturbations, and a combination of each. The calibration pose and
    zeroing are present in the data but not identified explicitly by an event.
  \item  Trials 6-8's force measurements are affected by the treadmill
    vibration mode mentioned in the equipment section and the force plate data
    should not be used. We include the trials because both the kinematic data
    is valid and it includes lateral perturbations in addition to the
    longitudinal.
  \item Trials 9, 10, and 11 used a slightly different event definition where the
    calibration poses were not explicitly tagged by an event, yet the protocol
    was identical to the following trials. The calibration pose will have to be
    determined manually.
  \item During trials 9-15, we used wooden blocks to fix the treadmill to the
    concrete floor to eliminate the treadmill's low vibration mode
    (approximately 5\si{\hertz}). But these blocks seem to have corrupted the
    force plate measurements by imposing frictional stresses on the system. The
    force plate measurements should not be used from these trials, but the
    marker data is fine.
  \item We did not record unloaded compensation trials for trials 9-15.
    Regardless, they would likely be useless due to the corruption from the
    wooden blocks.
  \item The ankle joint torques computed from subject 9's data in trials 25-27
    are abnormal and should be used with caution or not at all. We were not
    able to locate the source of the error, but it is likely related to the
    force calibration.
  \item Trials 37, 38, and 39 do not exist. The numbers were accidentally
    skipped.
\end{itemize}

\section*{Conclusion}
%
We have presented a rich and elaborate data set of motion and ground reaction
loads from human subjects during both normal walking and when recovering from
longitudinal perturbations. The raw data is provided for reuse with complete
meta data. In addition to the data, we provide software that can process the
data for both cleaning purposes and to produce typical sagittal plane gait
variables of interest. Among other uses, we believe the dataset is ideally
suited for control identification purposes. Many researchers are working on
mathematical models for control in gait and this dataset provides both a way to
validate these models and a source for generating them.

>>>>

% Editing notes
% - No tabs, just spaces!
% - No hanging whitespace.
% - Wrap lines to 79 characters.
% - Indentation should be two spaces per indent.
% - Only one space after periods.
% - Don't worry about formatting details, just get the content in place.

@@

\subsection*{Participants}
%
Fifteen able bodied subjects including four females and eleven males with an
average age of $24\pm4$ years, height of $1.75\pm0.09$ m, mass of $74\pm13$ kg
participated in the study. The study was approved by the Institutional Review
Board of Cleveland State University (\# 29904-VAN-HS) and written informed
consent was obtained from all participants. The data has been anonymized with
respect to the participants' identities and a unique identification number was
assigned to each subject. A selection of the meta data collected for each
subject is shown in Table~\ref{tab:subjects}.
%
\begin{table}
  \cprotect\caption{Information about the 15 study participants in order of
    collection date. The subjects are divided into those that were used for the
    protocol pilot trials, i.e. the first three, and those used for the final
    protocol. The final three columns provide the trial numbers associated with
    each nominal treadmill speed. The measured mass is computed from the mean
    total vertical ground reaction force just after the calibration pose event,
    if possible.  If the mass is reported without an accompanying standard
    deviation, it is the subject's self-reported mass.  Additional trials found
    in the data set with a subject identification number 0 are trials with no
    subject, i.e.  unloaded trials that can be used for inertial compensation
    purposes, and are not shown in the table. Generated by
    \verb|src/subject_table.py|.}
  \centering
  \small
  \input{tables/subjects.tex}
  \label{tab:subjects}
\end{table}

@@

Early on, we discovered that the factory setup of the R-Link treadmill had a
vibration mode as low as 5\si{\hertz} that was detectable in the force
measurements, likely due to the flexible undercarriage and pitch motion
mechanism. Trials 6--8 are affected by this vibration mode. During trials 9--15
the treadmill was stabilized with wooden blocks. During the remaining trials
($>$~15) the treadmill was stabilized with metal supports; both ones we
fabricated in-house and ones supplied by the vendor. These supports aimed to
improve the stiffness and frequency response of the force plate system. See the
Data Limitations Section for more details.

The acceleration of the treadmill base was measured during each trial by the
ADXL330 accelerometers placed at the four corners of the machine. These
accelerometers were intended to provide information for inertial compensation
purposes when the treadmill moved laterally or in pitch, but are extraneous for
trials greater than number 8 due to the treadmill being stabilized in those
degrees of freedom by the aforementioned supports.
%
\begin{figure}
  \centering
  \includegraphics{figures/treadmill.png}
  \caption{The treadmill with coordinate system, cameras (circled in orange),
    projection screen, and safety rope. The direction of travel is in the $-z$
    direction.}
  \label{fig:treadmill}
\end{figure}

@@

hair at the expense of accuracy.}, and a rock climbing harness. All 47 markers
were applied directly to the skin at the landmarks noted in

@@

  \cprotect\caption{Descriptions of the 47 subject markers used in this study.
    The ``Set'' column indicates whether the marker exists in the lower and/or
    full body marker set. The label column matches the column headers in the

@@

After obtaining informed consent and a briefing by the experimentalist on the
trial protocol, the subject followed the verbal instructions of the
experimentalist and the on-screen instructions from the video display. The
final protocol for a single trial was as follows:
%
\begin{enumerate}
  \item Subject stepped onto the treadmill and markers were identified with
    Cortex.
  \item The safety rope was attached loosely to the rock climbing harness such
    that no forces were acting on the subject during walking, but so that the
    harness would prevent a full fall.
  \item The subject started by stepping on sides of treadmill so that feet did
    not touch the force plates and the force plate signals are zeroed. This
    corresponds to the ``Force Plate Zeroing'' event.
  \item Once notified by the video display, the subject stood in the
    calibration pose: standing straight up, looking forward, arms out by their
    sides (approximately 45 degree abduction) and the event, ``Calibration
    Pose'', was manually recorded by the operator.
  \item A countdown to the first normal walking phase was displayed. At the end
    of the countdown the event ``First Normal Walking'' was recorded and the
    treadmill ramped up to the specified speed and the subject was instructed
    to walk normally, to focus on the ``endless'' road on the display, and not
    to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begun and was recorded as ``Longitudinal Perturbation''.
  \item After 8 minutes of walking under the influence of the perturbations,
    the second normal walking phase begun and was recorded as ``Second Normal
    Walking''.
  \item After 1 minute of normal walking, a countdown was shown on the display
    and the treadmill decelerated to a stop.
  \item The subject was instructed to step off of the force plates for 10
    seconds and the ``Unloaded End'' event was recorded.
  \item The subject could then take a rest break before each additional trial.
\end{enumerate}

\subsubsection*{Pilot Protocols}
%
Trials 3--15 were pilot tests for finalizing the protocol design an thus have
some slight variations with respect to the subsequent trials. We include these
trials due to the uniqueness of trials 6--8 and the fact that the kinematic
data is valid. We believe there may be useful analyses that only require the
kinematic data. Additional information needed to interpret the data in the
pilot trials can be found in the associated meta data files and the Data
Limitations Section of this paper.

Trials 6--8 use an early experimental protocol which divided the walking period
into three sections: no perturbation, longitudinal perturbation, and a
combination of longitudinal and lateral perturbation. The calibration pose and
zeroing events are present in the data but lumped into one event. These trials
only use the lower body marker set described in Table~\ref{tab:marker-labels}.
Additionally, there are five markers that have labels beginning with \verb|ROT|
that were attached to the treadmill base to capture the lateral motion. Trials
9--15 use the final protocol but have corrupt ground reaction loads due to the
wooden treadmill base stabilizers.

@@

In addition to the longitudinal perturbations, lateral perturbations were also
prescribed for a duration of four minutes in the pilot trials 3--8.
Figure~\ref{fig:lateral_perturbation} shows the effect of the lateral control
of the treadmill by plotting the commanded and measured treadmill positions.
These signals were generated in a similar manner using MATLAB and Simulink in
which a Gaussian white noise block that twice integrated to obtain the lateral
position. The signal was then high-pass filtered with a second-order
Butterworth filter to eliminate drift and then saturated so that the signal
remained within the 5~\si{\centi\meter} lateral range of the physical hardware.
The same perturbation signal was used for each of the three trials.
%
\begin{figure}
  \centering
  \includegraphics{figures/lateral_perturbation.pdf}
  \cprotect\caption{Commanded treadmill lateral deviation (red) and the
    recorded deviation (blue) for the trials 6, 7, and 8. Generated by
    \verb|src/lateral_perturbation_plot.m|.}
  \label{fig:lateral_perturbation}
\end{figure}

@@

\subsubsection*{mocap-xxx.txt}
%
The output from the D-Flow mocap module is stored in a tab separated value
(TSV) file named \verb+mocap-xxx.txt+ where \verb+xxx+ represents the trial id
number. The file contains a number of time series. The numerical values of the
time series are provided in decimal fixed point notation with 6 decimals of
precision, e.g. \verb|123456.123456|, regardless of the units. The first line
of the file holds the header. The header includes time stamp column, frame
number column, marker position columns, force plate force/moment columns, force
plate center of pressure columns, and other analog columns. The columns are
further described below:
%
\begin{description}
  \item[TimeStamp] The monotonically increasing computer clock time when D-Flow
    receives a frame from Cortex. These are recorded approximately at
    100~\si{\hertz} sampling rate and given in seconds.
  \item[FrameNumber] Monotonically increasing positive integers that correspond
    to each frame received from Cortex.
  \item[Marker Coordinates] Any column that ends in \verb+.PosX+, \verb+.PosY+,
    or \verb+.PosZ+ are marker coordinates expressed in Cortex's Cartesian
    reference frame. The prefixes match the marker labels given in
    Table~\ref{tab:marker-labels}. These values are in meters.
  \item[Ground Reaction Loads] There are three ground reaction forces and three
    ground reaction moments recorded by each of the two force plates in Newtons
    and Newton-Meters, respectively. The prefix for these columns is either
    \verb+FP1+ or \verb+FP2+ and represents either force plate 1 (left) or 2
    (right). The suffixes are either \verb+.For[XYZ]+, \verb+.Mom[XYZ]+ for the
    forces and moments, respectively. The force plate voltages are sampled at a
    much higher frequency than the cameras, but delivered at the Cortex camera
    sample rate, approximately 100~\si{\hertz}, through the D-Flow mocap
    module. A force/moment calibration matrix stored in Cortex converts the
    voltages to forces and moments before sending it to D-Flow. The software
    also computes the center of pressure from the forces, moments, and force
    plate dimensions. These have the same prefixes for the plate number, have
    the suffix \verb+.Cop[XYZ]+, and are given in meters.
  \item[Analog Channels] Several analog signals are recorded under column
    headers \verb+Channel[1-99].Anlg+. These correspond to analog signals
    sampled by Cortex and correspond to the 96 analog channels in the National
    Instruments USB-6255. The first twelve are the voltages from the
    force plate load cells. We also record the acceleration of 4 points on the
    treadmill base in analog channels 61--72 that were in place in case
    inertial compensation for the lateral treadmill movement was required.
\end{description}

@@

The \verb|DFlowData| class collects and stores all the raw data presented in
the previous section and applies several ``cleaning'' operations to transform
the data into a usable form. The cleaning process follows these steps:
%
\begin{enumerate}
  \item Load the meta data file into a Python dictionary.
  \item Load the D-Flow mocap module TSV file into Pandas \verb|DataFrame|.
  \item Relabel the column headers to more meaningful names if this is
    specified in the meta data.
  \item Optionally identify the missing values in the mocap marker data and
    replace them with \verb|numpy.nan|.
  \item Optionally interpolate the missing marker values and replaces them
    with interpolated estimates using a variety of interpolation methods.
  \item Load the D-Flow record module TSV file into a Pandas \verb|DataFrame|.
  \item Extract the events and create a dictionary mapping the event names in
    the meta data to the events detected in the record module file.
  \item Inertially compensate the ground reaction loads based on whether the
    meta data indicates there was treadmill motion.
  \item Merge the data from the mocap module and record module into one data
    frame at the maximum common constant sample rate.
\end{enumerate}

@@

\subsection*{Data Limitations}
%
The data is provided in good faith with great attention to detail but as with
all data there are anomalies that may affect the use and interpretation of
results emanating from the data. The following lists give various notes and
warnings about the data that should be taken into account before use.

\subsubsection*{All Trials}
%
\begin{itemize}
  \item Be sure to read the notes in each meta data file for details about
    possible anomalies in that particular trial. Things such as marker dropout,
    ghost markers, and marker movement are the more prominent notes. Details
    about variations in the equipment on the day of the trial are also
    mentioned.
  \item The subject identification number 0 represents the ``null subject'' and
    was used whenever data was collected from the system with no subject on the
    treadmill, for example during the trials that were intended to be used for
    inertial compensation purposes. These trials play through the exact
    protocol as those with a human subject and the matching trials are
    indicated in the meta data. Matching unloaded trials were recorded on the
    same day as the loaded trials and is noted in the
    \verb|trial:files:compensation| section of the meta data file. See
    Table~\ref{tab:comptrials} for a list of all the compensation trials.
  \item Trials 1 and 2 were not recorded as part of this study. Those trial
    identification numbers were reserved for early data exploration from data
    collected in other studies and are not included in this dataset.
  \item Trials 37, 38, and 39 do not exist. The numbers were accidentally
    skipped.
  \item The ankle joint torques computed from subject 9's data in trials 25--27
    are abnormal and should be used with caution or not at all. We were not
    able to locate the source of the error, but it is likely related to the
    force calibration.
\end{itemize}

\subsubsection*{Pilot Trials}
%
\begin{itemize}
  \item Subject 1 walked only at a single speed with three trial repetitions.
  \item Trials 6--8 included a calibration pose at the start of the trial but
    the event was not explicitly recorded. In those trials, the
    ``TreadmillPerturbation'' event marks the beginning of longitudinal
    perturbations and the ``Both'' event marks the beginning of combined
    longitudinal and lateral perturbations. The force plate zeroing at the end
    was also not explicitly recorded.
  \item Trials 6--8's force measurements are affected by the treadmill vibration
    mode mentioned in the equipment section and the force plate data should not
    be used.
  \item Trials 9--11 used a slightly different event definition where the
    calibration poses were not explicitly tagged by an event, yet the protocol
    was identical to the following trials. The calibration pose will have to be
    determined manually.
  \item During trials 9--15, we used wooden blocks to fix the treadmill to the
    concrete floor to eliminate the treadmill's low vibration mode. But these
    blocks seem to have corrupted the force plate measurements by imposing
    frictional stresses on the system. The force plate measurements should not
    be used from these trials.
  \item We did not record unloaded compensation trials for trials 9--15.
    Regardless, they would likely be useless due to the corruption from the
    wooden blocks and are not needed because the treadmill base is fixed.
\end{itemize}

\section*{Conclusion}
%
We have presented a rich and elaborate data set of motion and ground reaction
loads from human subjects during both normal walking and when recovering from
perturbations. The raw data is provided for reuse with complete meta data. In
addition to the data, we provide software that can process the data for both
cleaning purposes and to produce typical sagittal plane gait variables of
interest. Among other uses, we believe the dataset is ideally suited for
control identification purposes. Many researchers are working on mathematical
models for control in gait and this dataset provides both a way to validate
these models and a source for generating them.
